---
title: "MTHM506 - Statistical Data Modelling: Individual Project"
author: "James R Lewis"
date: "March 2025"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
    keep_tex: true
    includes:
      in_header: preamble.tex
    extra_dependencies: ["amsmath", "amssymb", "mathtools", "bm", "caption", "hyperref", "geometry", "setspace"]
  html_document:
    css: style.css
    toc: true
    toc_depth: 2
    number_sections: true
  word_document:
    toc: true
    toc_depth: 2
    number_sections: true
header-includes:
  - \usepackage{geometry}
  - \geometry{a4paper, margin=1in}
  - \usepackage{setspace}
  - \onehalfspacing
  - \usepackage{parskip}          # Adds space between paragraphs for readability
  - \setlength{\parindent}{0pt}   # Removes paragraph indentation for a cleaner look
  - \hypersetup{colorlinks=true, linkcolor=blue, urlcolor=blue, citecolor=blue} # Consistent link colours in PDF
knitr:
  opts_chunk:
    cache: false
    echo: true
    message: false
    warning: false
classoption: a4paper,11pt
fontsize: 11pt
link-citations: yes
---

```{r loading datasets and libraries, include=FALSE}
load("~/GitHub/university-projects/Statistical Data Modelling Projects - MTHM506/Coursework 2 - Project/datasets_project.RData")
TBdata <- TBdata

library(tidyverse)
library(ggplot2)
library(rmarkdown)
library(mgcv)

```

**Declaration of AI Assistance**: I have used OpenAI’s ChatGPT tool in creating this report.

AI-supported/AI-integrated use is permitted in this assessment. I acknowledge the following uses of GenAI tools in this assessment:

1.  I have used GenAI tools to check and debug my code.
2.  I have used GenAI tools to proofread and correct grammar or spelling errors.
3.  I have used GenAI tools to give me feedback on a draft.

I declare that I have referenced use of GenAI outputs within my assessment in line with the University referencing guidelines.

```{r setting plot theme, include = FALSE}
custom_theme <- theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),  # Centered title
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(size = 11), 
    panel.grid.major = element_line(color = "gray85"),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )
```

\newpage

# Introduction

-   **Objective:** Clearly state the aim of the analysis—to quantify TB risk across Brazil (2012-2014), identify socio-economic covariates affecting TB rates, and understand spatial, temporal, and spatio-temporal structures.

<!-- -->

-   **Relevance:** Discuss the public health importance and resource allocation implications.

<!-- -->

-   **Data Overview:** Briefly introduce the dataset, its source, and key variables (e.g., TB counts, socio-economic covariates, spatial and temporal identifiers).

# Exploratory Data Analysis (EDA)

-   **Descriptive Statistics:** Summarise key covariates and TB rates (use one table for this).



# Methodology

-   **GAM Framework:**

    -   Introduce GAMs as an extension of GLMs (refer to Topic 3 notes).

    -   Discuss the semi-parametric nature, choice of smoothing, and penalisation for smoothness.

    -   Explain the model structure, including:

        -   Response variable: TB rate (TB cases per unit population).
        - The offset

        -   Covariates: Socio-economic factors and spatial-temporal indicators.

        -   Link function and distribution assumption 
        - Why i went from NB to tw.
        

<!-- -->

-   **Model Formulation:** Provide the mathematical formulation of your GAMs, including how spatial and temporal structures are modelled.

Equations for Base tw model and temporal model
<!-- -->

-   **Justification of Choices:** Explain why GAMs are appropriate, especially for non-linear, spatial-temporal relationships.

# Model Fitting

-   **Model Development Steps:**

    -   **Initial Model:** Fit a base model with socio-economic covariates.

    -   **Spatial and Temporal Effects:** Extend the model to include smooth terms for spatial (latitude, longitude) and temporal (year) effects.

    -   **Spatio-Temporal Interaction:** Test for interactions if appropriate.

<!-- -->

-   **Model Selection and Evaluation:** Discuss criteria used for selecting the best model (e.g., AIC, GCV, residual analysis).

<!-- -->

-   **Diagnostics:** Present key diagnostics (e.g., residual plots, deviance explained).

# Results

-   **Summary of Findings:** Interpret the key model outputs:

    -   Which covariates significantly influence TB rates?

    -   How do spatial and temporal trends appear?

    -   Are there high-risk regions or periods for TB?

<!-- -->

-   **Visual Representation:**

    -   Include maps showing predicted TB risks.

    -   Temporal trend plots.

    -   Any significant interaction effects (if modelled).

<!-- -->

-   **Implications:** Discuss practical implications, particularly for resource allocation.

# Critial Review

-   **Model Limitations:** Reflect on any limitations (e.g., assumptions, data gaps, potential overfitting).

<!-- -->

-   **Alternative Approaches:** Briefly discuss other modelling approaches that could be considered in the future.

<!-- -->

-   **Uncertainty Discussion:** Highlight any uncertainties in parameter estimation, especially regarding smooth terms.

# Conclusion

-   **Summary of Key Insights:** Recap significant findings and their implications.

<!-- -->

-   **Policy Recommendations:** Suggest actionable steps for health authorities based on the analysis.

















## Appendix:

-   **Commented R Code:** Include all key code used for analysis and modelling.

<!-- -->

-   **Additional Figures/Tables:** Any extra plots or results not included in the main text.

<!-- -->

-   **Model Summaries:** Full output of final model summaries.

The aim of this project is to use this data set to quantify TB risk across Brazil over the 3 years, where risk is defined as the rate of TB cases per unit population.

**Tables**:

1.  Summary Stats Table (gt table, look at website) done
2.  Model Comparison (Tweedie vs nb) done
3.  Table for regions iwth highest rate of TB, and infers from this

**Plots**:

1.  Residual Diagnostic plots: Residual vs fitted QQ plots and 1 other? Influece plot?

2.  Covariate plot 1-8 covaraite impact on rate of TB per unit pop. Compare the raw mean to predicted mean.

3.  Spatial structure explaining risk: plot.map for this. General plot without temporal noise

4.  Temporal structure explaining risk: Covariate charts/significance for each year

5.  temporal-Spatial structure explaining risk: plot.map for this. 3 of them, Look at differences

6.  

The health authorities want to allocate resources for hospitals to cope with the TB cases , so they would like to know if there are regions where the rate of TB per unit population is high and where you would recommend allocating these resources.

```{r}
 library(fields)
 library(maps)
 library(sp)
 # PLotting map of cases
 plot.map(TBdata$TB[TBdata$Year==2014],n.levels=7,main="TB counts for 2014")
```

Structure:\
\
\## **1. Exploratory Data Analysis (EDA)**

-   **Descriptive Statistics:** Summary of key covariates and TB case rates (per unit population).

Write about Table 1

And implications going into this.

## **2. Initial Model Fitting (GAMs)**

-   **Base Model:** Fit an initial GAM including socio-economic covariates.

-   **Model Structure Explanation:** Detail the mathematical formulation and rationale for including each covariate.

-   **Model Summary:** Present key outputs (e.g., significant predictors, smooth terms).

We will model **TB case counts** $Y_{i}​$ as a function of socio-economic covariates and spatial-temporal effects using **Generalized Additive Models (GAMs)**. Since TB cases are **count data**, they follow a **Poisson or Negative Binomial (NB) distribution**, which makes the following GAM structure appropriate:

## Model Specification

We assume that the observed TB case counts $Y_i$ follow either a Poisson or a Negative Binomial distribution:

$$
Y_i \sim \text{Poisson}(\mu_i) \quad \text{or} \quad Y_i \sim \text{NB}(\mu_i, \theta)
$$

where the expected mean $\mu_i$ is modeled as:

$$
\log(\mu_i) = \beta_0 + f_1(\text{Indigenous}_i) + f_2(\text{Illiteracy}_i) + \dots + f_p(\text{Timeliness}_i) + f_{\text{spatial}}(\text{lon}_i, \text{lat}_i) + f_{\text{temporal}}(\text{Year}_i) + \log(\text{Population}_i)
$$

Here:

\- $f_j(\cdot)$ represents smooth functions capturing non-linear effects of the socio-economic covariates.

\- $f_{\text{spatial}}(\text{lon}, \text{lat})$ models spatial variability.

\- $f_{\text{temporal}}(\text{Year})$ accounts for temporal trends.

\- The $\log(\text{Population}_i)$ term acts as an offset to adjust for population size.

This formulation allows us to estimate the impact of socio-economic factors on TB case counts while incorporating spatial and temporal structures in the data.

### **Choice of Distribution: Poisson vs. Negative Binomial**

#### **Limitations of the Poisson Model**

The Poisson model assumes that the mean and variance of the response variable are equal:

$$
E[Y] = \text{Var}(Y)
$$

However, this assumption is often violated in real-world count data due to **overdispersion**, where the variance exceeds the mean. Overdispersion can lead to:

\- **Inflated test statistics**, increasing the risk of Type I errors (false positives).

\
- **Misleading inferences**, as confidence intervals may be too narrow.

#### **Evidence of Overdispersion**

The computed dispersion statistic for the TB data is **2223.535**, which is significantly greater than 1. This confirms severe overdispersion, violating the equidispersion assumption of the Poisson model:

$$
\text{Var}(Y) \gg E[Y]
$$

Using a Poisson model under such conditions would underestimate variability, leading to unreliable statistical conclusions.

#### **Why the Negative Binomial Model is More Appropriate**

The **Negative Binomial (NB) model** extends the Poisson distribution by introducing an **overdispersion parameter** $\theta$, which adjusts the variance function to account for excess dispersion:

$$
\text{Var}(Y_i) = \mu_i + \frac{\mu_i^2}{\theta}
$$

where:

\- $\mu_i$ is the expected number of TB cases for region $i$, - $\theta$ controls the degree of overdispersion, allowing for greater variability than the Poisson model.

Given the extreme overdispersion in the TB data, the Negative Binomial model is the **statistically robust choice**, as it:

\- **Relaxes the restrictive assumption of equal mean-variance** while remaining within the exponential family framework.

\- **Provides more reliable parameter estimates** by accounting for unobserved heterogeneity across microregions.

\- **Prevents misleading inference** by correcting for variance inflation.

#### **Justification for Directly Using a Negative Binomial GAM**

Rather than initially fitting a Poisson GAM and later diagnosing overdispersion, we **directly proceed with a Negative Binomial GAM**. This ensures:

\- **Model validity**, by selecting an appropriate variance structure from the outset.

\- **Robust inference**, reducing the risk of biased estimates and misleading conclusions.

\- **Improved accuracy**, as the model better captures the heterogeneity in TB case counts across microregions.

By making this adjustment, we ensure a more reliable and interpretable analysis of TB risk in Brazil.

## 3. Offset for Population

Since TB cases depend on the **population size** in each region, we introduce an **offset** to model TB risk per capita:

$$
\log(\mu_i) = \text{linear predictor} + \log(\text{Population}_i)
$$

This ensures that the response variable is scaled appropriately, effectively modeling the **rate of TB cases per 100,000 people** rather than just raw counts.

## **Negative Binomial GAM for TB Case Counts**

Given the substantial **overdispersion** in the TB case counts, we proceed with a **Negative Binomial (NB) GAM**. The model accounts for **spatial, temporal, and socio-economic effects** while incorporating an **offset term** to model TB risk per unit population.

### **Mathematical Formulation**

The **Negative Binomial GAM** assumes that the response variable $Y_i$ (TB case count in region $i$) follows a **Negative Binomial distribution**:

$$
Y_i \sim \text{NB}(\mu_i, \theta)
$$

where:

\- $\mu_i$ is the expected number of TB cases in region $i$. - $\theta$ is the **overdispersion parameter**.

The **log link function** ensures a **multiplicative relationship** between the covariates and the expected count:

$$
\log(\mu_i) = \beta_0 + f_1(\text{Indigenous}_i) + f_2(\text{Illiteracy}_i) + \dots + f_p(\text{Timeliness}_i) + f_{\text{spatial}}(\text{lon}_i, \text{lat}_i) + f_{\text{temporal}}(\text{Year}_i) + \log(\text{Population}_i)
$$

The **offset term** $\log(\text{Population}_i)$ ensures that the model accounts for diffent population sizes.

------------------------------------------------------------------------

## **Transforming Predictions to TB Incidence Rates**

Since the **Negative Binomial GAM** inherently predicts **TB case counts**, we manually convert the **fitted values** into **TB incidence rates per 100,000 population** using:

$$
\text{Predicted TB Rate} = \left( \frac{\text{Predicted TB Cases}}{\text{Population}} \right) \times 100000
$$

This transformation ensures that the estimated values are correctly **interpreted as TB incidence rates** rather than raw counts.

\########################

## **Understanding the Model Output**

Even though we include an **offset term** $\log(\text{Population})$ in the model to account for different **population sizes**, the model still fundamentally **predicts TB case counts**.

### **Mathematical Formulation**

The mathematical form of the model is:

$$
\log(E[Y_i]) = \beta_0 + f_1(\text{Indigenous}_i) + f_2(\text{Illiteracy}_i) + \dots + f_p(\text{Timeliness}_i) + f_{\text{spatial}}(\text{lon}_i, \text{lat}_i) + f_{\text{temporal}}(\text{Year}_i) + \log(\text{Population}_i)
$$

where: - $E[Y_i]$ = expected **TB case count** in microregion $i$, - $f_j(\cdot)$ = **smooth functions** of socio-economic covariates, - $f_{\text{spatial}}(\cdot)$ = **spatial effect**, - $f_{\text{temporal}}(\cdot)$ = **temporal effect**, - $\log(\text{Population}_i)$ = **offset term** to adjust for population differences.

Since the model uses a **log link function**, the expected case count is given by:

$$
E[Y_i] = \text{Population}_i \times e^{\text{Linear Predictor}}
$$

### **Implication**

-   The model’s predictions are in terms of **TB case counts**, adjusted for population.
-   It **does not directly predict TB incidence rates**.
-   To obtain **TB rates**, we must **manually convert** the predicted case counts.

------------------------------------------------------------------------

## **2. Why Does This Happen Despite the Offset?**

The **offset does not change** the response variable from **counts to rates**. Instead, it ensures that the estimated **case counts are proportional to the population size**.

### **Key Insights**

-   If two regions have the **same covariate values** but **different populations**, the model will predict **higher TB case counts** for the region with the larger population.
-   The offset ensures that **TB risk per person is correctly estimated**, but the raw predictions remain in **case counts**.

Thus, while the model **adjusts for population**, it does not inherently provide **incidence rates**, requiring us to **explicitly transform** the predictions.

\##########################

### **Fitting the Model in R**

We use the `mgcv` package to fit a **Negative Binomial GAM**, specifying the **offset for population size**:

## **3. Model Refinement and Improvement**

-   **Add Complexity:** Introduce spatial and temporal smooths to capture unexplained variation.

-   Explain why some terms are included and some not. why some are formatted as they as.. Illitercy seems to have no significant correlation as seen from a plot.

-   **Interaction Terms:** If necessary, test for spatio-temporal interactions.

-   **Summary Table:** Present a table comparing models (AIC, deviance explained, etc.) to show the progression and improvements.

## **4. Model Diagnostics**

-   **Residual Diagnostics:** Plot residuals to assess model fit (QQ plots, residual vs fitted, etc.).

-   **Check Assumptions:** Comment on model assumptions like distributional assumptions, independence, and overfitting.

## **5. Covariate Effect Plots** (Already created)

-   **Effect Plots for Covariates:** Show smooth effect plots (e.g., `plot.gam` in R) for key socio-economic variables to visualise their impact on TB risk.

-   **Interpretation:** Discuss how each covariate affects TB risk and the strength of their influence.

## **6. Spatial and Temporal Structure Analysis**

-   **Spatial Smoother Visualization:** Plot the estimated spatial effects to identify regions with high or low TB risk.

-   **Temporal Smoother Visualization:** Show how TB rates have evolved over time.

## **7. Geospatial Plots**

-   **Risk Maps:** Use the `plot.map` function to produce maps of predicted TB risks across microregions.

-   **Highlight High-Risk Areas:** Identify and discuss regions with consistently higher TB rates.

## **9. Model Comparison**

-   **Performance Summary:** Compare final models based on criteria like AIC, deviance explained, and residual diagnostics.

-   **Final Model Selection:** Clearly state which model is preferred and why.

## **10. Conclusion and Recommendations**

-   **Key Insights:** Summarise the main findings regarding socio-economic influences, spatial-temporal risks, and high-risk areas.

-   **Policy Implications:** Discuss recommendations for health authorities, such as where to allocate resources.
